The hydra and the rooster

### 2020-07-13 latex/chapter-hydras.tex
more text on omega plus opmega

\subsection{Some failed attempts to prove termination}
The class \texttt{Hvariant} is parameterized by a well-founded order $<$ on some type $A$, a class of battles $B$ and a measure $m$. When $<$ or $m$ are 
badly given for a given class $B$, one may fail to find a variant.

In this section, we present some naive attempts, where we fail to build proof of termination of free battles.

\subsubsection{Using the Hydra's height as a variant}
A first plan to prove termination of all hydra battles is to use a simple
measure that maps any hydra to a natural number. For instance, let us check whether 
 the function \emph{height} could be such  a variant.

Unfortunately, that does not work.
Please consider the hydras of Fig.~\vref{fig:h-omega2-small} and
\ref{fig:h-omega-omega}. The former can be transformed into the latter in one round, but their heights are equal.

@@

\subsubsection{Using any variant defined on \texorpdfstring{nat}{nat}}
\label{omega-case}
We could imagine that our previous failure is due to the choice of \texttt{height} as a measure, and that a more complex function would work.
In fact, we can prove that 
\emph{no} instance of class \texttt{WfVariant round Peano.lt $m$} can be build, where
$m$ is \emph{any} function of type \texttt{Hydra $\arrow$ nat}.

@@

\subsubsection{Lexicographic order on \texorpdfstring{nat*nat}{nat*nat}}
\label{omega2-case}

We  prove now that even the type \texttt{nat * nat}, provided with the lexicographic product of \texttt{(nat,<)} by itself is too simple for proving the termination of all hydra battles. This impossibility result will prevent us from considering measures like the following one:

  \begin{Coqsrc}
  Let m h = (height h, hsize h).
  \end{Coqsrc}

The proof we are going to develop has exactly the same structure as the previous one. Nevertheless, the proof of technical  lemmas is a little more complex, due to 
 the structure of the lexicographic order on $\mathbb{N}\times\mathbb{N}$. 
Consider for instance that there exists an infinite number of pairs between
$(1,0)$ and $(2,0)$.

\begin{remark}
  The order structure we consider in this section is also known as the ordinal
  $\omega^2$. We identify any pair $(i,j)\in \mathbb{N}\times\mathbb{N}$ with the ordinal $\omega\times i + j$. Thus the three kinds of ordinals in $\omega^2$
  are represented as follows:
  \begin{description}
  \item[null ordinal] : the pair $(0,0)$
  \item[successor ordinal] : any pair $(i,j)$ where $j>0$
  \item[limit ordinal] : any pair $(i,0)$ where $i>0$
  \end{description}
\end{remark}

The detailed  proof script is in the file \url{../src/html/hydras.Hydra.Omega2_Small.html}.

\subsubsection{Preliminaries}
Let us assume there is a variant from \texttt{Hydra} into \texttt{nat*nat} (with the
   lexicographic ordering)  for proving the   termination of all hydra battles.

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Omega2_Small.html}{Hydra.Omega2\_Small}}

\begin{Coqsrc}
  Section Impossibility_Proof.

  Let t := (nat * nat)%type.

  (** non-dependent lexicographic strict ordering on nat*nat *)

  Let lt2 : relation t := lexico Peano.lt Peano.lt.

  Infix "<" := lt2.

  (** reflexive closure of lt2 *)
  Let le2 := clos_refl _ lt2.
  Infix "<=" := le2.

  Variable m : Hydra -> t.

  Context (Hvar : Hvariant lt2_wf free m).
\end{Coqsrc}

Let us follow the same pattern as in Sect.~\ref{omega-case}.
First, we define an injection from type \texttt{t} into \texttt{Hydra}.
We associate with any pair $(i,j)$ the hydra with $i$ branches of length $2$ and
$j$ branches of length $1$.

%% revenir ici

\vspace{4pt}
\emph{From Module ~\href{../src/html/hydras.Hydra.Omega2_Small.html\#iota}{Hydra.Omega2\_Small}}

\begin{Coqsrc}
  Let iota (p: t) := 
    node (hcons_mult (hyd1 head) (fst p)
                     (hcons_mult head (snd p) hnil)).
\end{Coqsrc}

For instance, Figure~\vref{fig:essai2} shows the hydra associated to the pair
$(3,5)$.

\begin{figure}[htb]
\centering
\begin{tikzpicture}[very thick, scale=0.4]
\node (foot) at (6,0) {$\bullet$};
\node (N1) at (1,3) {$\bullet$};
\node (N2) at (3,3) {$\bullet$};
\node (N3) at (5,3) {$\bullet$};
\node (N4) at (8,3) {$\Smiley[2][green]$};
\node (N5) at (11,3) {$\Smiley[2][green]$};
\node (N6) at (14,3) {$\Smiley[2][green]$};
\node (N7) at (17,3){$\Smiley[2][green]$};
\node (N8) at (20,3){$\Smiley[2][green]$};
\node  (N9) at (0,5) {$\Smiley[2][green]$};
\node (N10) at (2,5) {$\Smiley[2][green]$};
\node (N11) at (4,5) {$\Smiley[2][green]$};
\draw (foot) to [bend left=10] (N1);
\draw (foot) -- (N2);
\draw (foot) -- (N3);
\draw (foot) -- (N4);
\draw (foot) -- (N5);
\draw (foot) -- (N6);
\draw (foot) to [bend right=10] (N7);
\draw (foot) to [bend right=15] (N8);
\draw (N1) to [bend left=10] (N9);
\draw (N2) -- (N10);
\draw (N3) -- (N11);
\end{tikzpicture}
\caption{\label{fig:essai2}
The hydra $\iota(3,5)$}
\end{figure}

Like in Sect.~\ref{omega-case}, we build a hydra out of the range of \texttt{iota} (represented in Fig.~\vref{fig:h-omega2-small}).

\begin{figure}[htb]
\centering
\begin{tikzpicture}[very thick, scale=0.5]
\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (3,4) {$\Smiley[2][green]$};
\node (N3) at (1,4) {$\Smiley[2][green]$};
\draw (foot) -- (N1);
\draw (N1) to [bend right =15] (N2);
\draw (N1) to  [bend left=15](N3);
\end{tikzpicture}
\caption{\label{fig:h-omega2-small}}
 The hydra \texttt{big\_h}.
\end{figure}
\begin{Coqsrc}
   Let big_h := hyd1 (hyd2 head head).  
 \end{Coqsrc}

 In a second step, we build a ``smaller'' hydra.

\begin{Coqsrc}
   Let small_h := iota (m big_h).
\end{Coqsrc}

Like in Sect.~\ref{omega-case}, we prove the double inequality \texttt{m big\_h <= m small\_h < m big\_h}, which is impossible.

\subsubsection{Proof of the inequality \texttt{m small\_h < m big\_h}}

For proving the inequality  \texttt{m\_lt: m small\_h < m big\_h}, it suffices to
build a fight transforming \texttt{big\_h} into \texttt{small\_h}.

First we prove that \texttt{small\_h} is reachable from \texttt{big\_h} in one or two steps. Let us decompose \texttt{m big\_h}
into the pair $(i,j)$.
If $j=0$, then one round suffices to transform \texttt{big\_h} into $\iota(i,j)$.
If $j>0$, then a first round transforms \texttt{big\_h} into $\iota(i+1,0)$ and a second round into $\iota(i,j)$. So, we have the folowing result.

\begin{Coqsrc}
 Lemma big_to_small: big_h -+-> small_h.
\end{Coqsrc}

Since $m$ is a variant, we infer the following inequality:

\begin{Coqsrc}
Corollary m_lt : m small_h < m big_h.
\end{Coqsrc}

\subsubsection{Proof of the inequality \texttt{m big\_h <= m small\_h} }

The proof of the inequality \texttt{m big\_h <= m small\_h} is quite more complex than in Sect~\ref{omega-case}.  If we consider some pair $(i,j)$, where $i>0$, there exists an infinite number of
pairs stricly less than $(i,j)$, and there exists an infinite number of battles that start from
$\iota(i,j)$. In effect, at any configuration $\iota(k,0)$, the hydra can freely chose any replication number. Intuitively, the measure of such a hydra must be large enough for taking into account
all the possible battles issued from that hydra.
Let us give more technical details.

\begin{itemize}
\item The proof of the lemma \texttt{m\_ge : forall p : t,   p <= m (iota p)} uses well-founded induction on $p$, and not structural induction on natural numbers

\item For any pair $p$, we have to distinguish between three cases, according to the value of $p$'s components.
  \begin{itemize}
  \item $p=(0,0)$
  \item $p=(i,0)$, where $i>0$\,: $p$ corresponds to a limit ordinal
  \item $p=(i,j)$, where $j>0$\,: $p$ is the successor of $(i,j-1)$.
  \end{itemize}
\end{itemize}

Before starting the proof, we have to express the notion of \emph{limit} in terms of
least upper bounds, through the following logical equivalence.

\vspace{4pt}
\emph{From Module ~\href{../src/html/hydras.Hydra.Omega2_Small.html\#limit_is_lub}{Hydra.Omega2\_Small.v}}.

\begin{Coqsrc}
Lemma limit_is_lub : forall i p, 
   (forall j, (i,j) < p) <-> (S i, 0) <= p.  
\end{Coqsrc}

Let us define the notion of elementary ``step'' of decreasing sequences in
\texttt{t}

\begin{Coqsrc}
Inductive step : t -> t -> Prop :=
| succ_step : forall i j,  step (i, S j) (i, j)
| limit_step : forall i j, step (S i, 0) (i, j).
\end{Coqsrc}

The following lemma establishes a correspondance between the relation
\texttt{step} and hydra fights.

\begin{Coqsrc}
Lemma step_to_fight : forall p q, step p q -> iota p -+-> iota q.
\end{Coqsrc}

\index{Maths!Transfinite induction}

Thus, starting from any inequality $q < p$ on type \texttt{t}, we can build 
by transfinite induction over \texttt{p} a fight 
that transforms the hydra $\iota(p)$ into $\iota(q)$.

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Omega2_Small.html\#m_ge}{Hydra.Omega2\_Small}}

\begin{Coqsrc}
Lemma m_ge : forall p : t,   p <= m (iota p).
Proof.
  intro p ; pattern p;
    apply  well_founded_induction with 
               (R := lt2) (1:= wf_lexico lt_wf lt_wf);
     intros (i,j) IHij (* rest of proof skipped *)
\end{Coqsrc}

\begin{Coqanswer}
  i, j : nat
  IHij : forall y : t, y < (i, j) -> y <= m (iota y)
  ============================
  (i, j) <= m (iota (i, j)) 
\end{Coqanswer}

Then we have  three cases to consider, according to the values of $i$ and $j$.
\begin{itemize}
\item If $p=(0,0)$ then obviously, $\iota(p)\geq p = (0,0)$
\item If  $p=(i+1,0)$ for some $i\in\mathbb{N}$, we
 remark  that $p$ is strictly greater than any pair $ (i, j)$, where $j$ 
is any natural number.

Applying the battle rules, for any $j$, we have $\iota(i+1,j)  {\round} \iota(i, j) $, thus $m(\iota(p)) > m(\iota(i,j)$ since  $m$ is assumed to be a variant.

Applying the induction hypothesis, we get the inequality
 $ m(\iota(i,j)) \geq (i,j)$ for any $j$.

Thus, $m(\iota(p)) > (i,j)$ for any $j$.
Applying the lemma \texttt{limit\_is\_lub}, we get  the inequality
$\iota(i+1,0)\geq (i+1,0)$

\item If $p=(i,j+1)$ with $j\in\mathbb{N}$, we have  $\iota(p)  {\round} \iota(i, j) $,
hence $m(\iota(p))> m(\iota(i,j)) \geq (i,j)$, thus $m(\iota(p))\geq (i,j+1)=p$

\end{itemize}

\subsubsection{End of the proof}

Since $<$ is a strict order (irreflexive  and transitive) on \texttt{nat*nat}, we can conclude that there is no
variant for termination on the lexicographic square of $(\mathbb{N},<)$.

\begin{enumerate}
\item From \texttt{m\_lt}, we infer the strict inequality 
\texttt{m small\_h < m big\_h}
\item from \texttt{m\_ge}, we get \texttt{m big\_h <= m (iota (m big\_h)) = m small\_h} 
\end{enumerate}

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Omega2_Small.html}{Hydra.Omega2\_Small}}

\begin{Coqsrc}
Theorem Impossible : False.
Proof.
  destruct (StrictOrder_Irreflexive  (m big_h)).
  apply le2_lt2_trans with (m small_h).
  -  unfold small_h;  apply m_ge.
  -  apply m_lt. 
Qed.

\begin{exercise}
Prove that there exists no variant $m$ from \texttt{Hydra} into \texttt{nat*nat} for proving
    the  termination of all \emph{standard} battles.
\end{exercise}

\begin{remark}
In Chapter~\ref{ks-chapter}, we will prove a much more general  theorem
than the two previous results about $\mathbb{N}$ and $\mathbb{N}\times\mathbb{N}$. The proof of that general  result will share the same structure, but will 
require a lot of technical results. 
\end{remark}

\begin{exercise}

\label{sec:orgheadline63}
Write \emph{direct} proofs ({i.e.},  without applying the result and tools of Chap.~\ref{ks-chapter}) that the following data structures  are too simple for defining a variant for any hydra battle.

\begin{itemize}
\item  $\omega^n$ : the set of all $n$-uples of natural numbers, ordered  by 
  lexicographic ordering
\item  $\omega^\omega$: the set of all decreasing sequences (with respect to $\le$)  of natural numbers, ordered by lexicographic ordering on lists.

For instance , the sequence $\langle 4,4,2 \rangle$ is strictly greater than 
$\langle 4,3,3,3,3,3,3,2,2,2 \rangle$.
\end{itemize}

>>>>

\subsection{A first theorem of impossibility}

The class \texttt{Hvariant} is parameterized by a well-founded order $<$ on some type $A$, a class of battles $B$ and a measure $m$. When $<$ or $m$ are 
badly given for a given class $B$, one may fail to find a variant.

We show that this is the case of the type \texttt{nat} of Peano natural numbers, provided with the usual strict order $<$.

\subsubsection{Using the Hydra's height as a variant}
Trivially, a measure like the height will not work for building a variant.

Please consider the hydras of Fig.~\vref{fig:h-omega2-small} and
\ref{fig:h-omega-omega}. The former can be transformed into the latter in one round, but their heights are equal.

@@

\subsubsection{Using any variant defined on \texorpdfstring{nat}{nat}}
\label{sect:omega-case}
We could imagine that our previous failure is due to the choice of \texttt{height} as a measure, and that a more complex function would work.
In fact, we can prove that 
\emph{no} instance of class \texttt{WfVariant round Peano.lt $m$} can be build, where
$m$ is \emph{any} function of type \texttt{Hydra $\arrow$ nat}.

@@

\subsubsection{Conclusion/Teaser}

In order to build a variant for proving the termination of all hydra battles, we need to consider order structures more complex than the usual order on type \texttt{nat}. The next chapter will present the notion of \emph{ordinal notation} from the point view of the \coq{} user: inductive types, functions, type classes, etc. 
The ordinal $\epsilon_0$, the first one allowing us to build our desired variant, will be studied in detail in 
Chapter~\ref{chap:T1}.

### 2020-07-13 latex/chapter-ordinals.tex
renaming zero_succ_limit-dec
more text on omega plus opmega

@@

@@

\end{remark}
\subsubsection{Definition of \texorpdfstring{$\omega$}{omega}}

@@

\subsection{Limit and successor ordinals}

@@

We say also that $\omega$ is the \emph{limit} of the sequence $\textit{fin}\;i, i\in\mathbb{N}$, according to the followi,g definition (in~\href{../src/html/hydras.Prelude.Ordinal_generic.html}{Prelude.Ordinal\_generic}).
$\alpha$ is the $\omega$-limit of the sequence $s$ if $\alpha$ is [strictly] larger than all the $s(i)$, and for any
$\beta<\alpha$, $\beta$ is dimonated by at least an element of  the sequence $s$ (in fact $\beta$ is dominated by an infinite number of elements of the sequence (\emph{exercise!})).

\begin{Coqsrc}
Definition  Omega_limit
            {A:Type}{lt : relation A}
           {sto : StrictOrder lt} (s: nat -> A) (alpha:A)  :=
  (forall i: nat, lt (s i) alpha) /\
  (forall beta, lt beta  alpha -> exists i:nat, lt beta (s i)).

Lemma omega_is_limit:
   Omega_limit fin omega.
\end{Coqsrc}

\begin{exercise}
Prove that $\omega$ is the unique   limit ordinal \emph{(in this precise notation system! In larger segments of ordinal, there will be \emph{a lot} of limit ordinals}).
\end{exercise}

Let $\alpha$ and $\beta$ be some ordinals, we say that \emph{$\alpha$ is an immediate predecessor of $\beta$}
(or that \emph{$\beta$ is a  successor of $\alpha$ }) if $\alpha<\$beta$ and there is no ordinal $\gamma$ between $\alpha$ and $\gamma$.

\noindent\emph{From Module~\href{../src/html/hydras.Prelude.Ordinal_generic.html}{Prelude.Ordinal\_generic}}

@@

\begin{itemize}
\item $\omega^n$ (for some  integer  $n\geq 2$) : the type of $n$-uples of natural numbers, with the lexicographic product of $n$ copies of \texttt{Nat.lt}.
\item  $\omega^\omega$: the set of nonincreasing sequences of natural numbers, lexicographically ordered (also the set of finite multisets of natural numbers).
\item $\epsilon_0$: The set of terms in Cantor normal form (see Chap.~\ref{chap:T1}).
\item $\Gamma_0$: The set of terms in Veblen normal form.
\end{itemize}

>>>>

\centerline{\LARGE {\bf work in progress!}}

@@

\begin{remark}
It is important to keep in mind  that the integer $n$ is not an ``element'' of \texttt{FinOrd $n$}. In set-theoretic presentations of ordinals, the set associated with the ordinal $n$ is $\{0,1,\dots,n-1\}$. 
In our formalization, the interpretation of an ordinal as a set is realized by the following definition
(in ~\href{../src/html/hydras.Prelude.Ordinal_generic.html}{Prelude.Ordinal\_generic}).

\begin{Coqsrc}
Definition bigO `{nA : @OrdinalNotation A ltA stoA compareA}
           (a: A) : Ensemble A :=
  fun x: A => ltA x a.
\end{Coqsrc}
\end{remark}

@@

\subsubsection{Definition of \texorpdfstring{$\omega$}{omega}}

@@

\subsubsection{Building an instance of OrdinalNotation}

The class \texttt{OrdinalNotation} is parameterized by the strictorder $<$ and a correct comparison function.

First, we prove that \texttt{lt} is an instance of \texttt{StrictOrder}.

\begin{Coqsrc}
Instance lt_strorder : StrictOrder lt.
Qed.
\end{Coqsrc}

Then, we define a comparison function by pattern matching over the sum type $t$.

\begin{Coqsrc}
Definition compare (alpha beta: t) : comparison :=
   match alpha, beta with
     inl _, inr _ => Lt
   | inl n, inl p | inr n, inr p => Nat.compare n p
   | inr _, inl _ => Gt
  end.
\end{Coqsrc}

\begin{Coqsrc}
Compute compare 7 omega.
\end{Coqsrc}

\begin{Coqanswer}
= Lt : comparison
\end{Coqanswer}

\begin{Coqsrc}
Compute compare  (inr 2) 42.
\end{Coqsrc}

\begin{Coqanswer}
= Gt : comparison
\end{Coqanswer}

\begin{Coqsrc}
Compute compare (inr 0) omega.  
\end{Coqsrc}

 \begin{Coqanswer}
= Eq : comparison
\end{Coqanswer}

Correctness of the function \texttt{compare} is asserted by the following lemma:

\begin{Coqsrc}
Lemma compare_correct alpha beta :
    CompareSpec (alpha = beta) (lt alpha beta) (lt beta alpha)
              (compare alpha beta).
\end{Coqsrc}

%%%% ICI %%%%

\subsubsection{Limit and successor ordinals}

The notions of limit and successor are defined in the module 
\href{../src/html/hydras.Prelude.Limits_and_Successors.html}{Prelude.Limits\_and\_Successors} for any instance of
\texttt{StrictOrder}.

@@

Let $<$ be a strict order on a type $A$ .
We say $x:A$ is an  $\omega$-\emph{limit} of  a sequence $s(i), i\in\mathbb{N}$, if
 if $x$ is [strictly] larger than all the $s(i)$, and for any
$y<x$, $y$ is dominated by at least an element of  the sequence $s$.

\vspace{4pt}

\noindent\emph{From Module~\href{../src/html/hydras.Prelude.Limits_and_Successors.html}{Prelude.Limits\_and\_Successors}}

\begin{Coqsrc}
Definition  Omega_limit
            {A:Type}{lt : relation A}
           {sto : StrictOrder lt} (s: nat -> A) (x:A)  :=
  (forall i: nat, lt (s i) x) /\
  (forall y, lt y  x -> exists i:nat, lt y (s i)).

Definition Limit   {A:Type}{lt : relation A}
           {sto : StrictOrder lt}  (x:A)  :=
  exists s : nat -> A, Omega_limit s x.
 \end{Coqsrc}

Let $x$ and $y$ be two terms of type $A$, we say that \emph{$x$ is an immediate predecessor of $y$}
(or that \emph{$y$ is a  successor of $x$ }) if $x<y$ and there is no $z$ between $x$ and $y$.

@@

For instance, 
we prove that $\omega$ is the limit of the sequence  of finite ordinals.

  \begin{Coqsrc}
Lemma omega_is_limit: Omega_limit fin omega.
\end{Coqsrc}

\begin{exercise}
Prove that $\omega$ is the unique   limit ordinal within the notation system for $\omega+\omega$
\emph{(in this precise notation system! In larger segments of 
ordinals, there will be \emph{a lot} of limit ordinals}).
\end{exercise}

\subsection{Computing in \texorpdfstring{$\omega+\omega$}{omega+omega}}

The simple data type we use for representing the ordinals less than $\omega+\omega$ allows us to compute successors, and to decide properties like \texttt{Successor} or \texttt{Limit}.

\begin{todo}
trichotomy zero/succ/limit
boolean functions limitb, succb

\end{todo}
\begin{itemize}
\item $\omega^n$ (for some  integer  $n\geq 2$) : the type of $n$-uples of natural numbers, with the lexicographic product of $n$ copies of \texttt{Nat.lt}.
\item  $\omega^\omega$: the set of nonincreasing sequences of natural numbers, lexicographically ordered (also the set of finite multisets of natural numbers).
\item $\epsilon_0$: The set of terms in Cantor normal form (see Chap.~\ref{chap:T1}).
\item $\Gamma_0$: The set of terms in Veblen normal form.
\end{itemize}

\begin{todo}
The following section has moved from chaper on hydras.
Adapt the redaction to the $\omega^2$ ordinl segment.
\end{todo}

\subsubsection{Lexicographic order on \texorpdfstring{nat*nat}{nat*nat}}
\label{omega2-case}

We  prove now that even the type \texttt{nat * nat}, provided with the lexicographic product of \texttt{(nat,<)} by itself is too simple for proving the termination of all hydra battles. This impossibility result will prevent us from considering measures like the following one:

  \begin{Coqsrc}
  Let m h = (height h, hsize h).
  \end{Coqsrc}

The proof we are going to develop has exactly the same structure as in Section~\ref{sect:omega:case}

previous one. Nevertheless, the proof of technical  lemmas is a little more complex, due to 
 the structure of the lexicographic order on $\mathbb{N}\times\mathbb{N}$. 
Consider for instance that there exists an infinite number of pairs between
$(1,0)$ and $(2,0)$.

\begin{remark}
  The order structure we consider in this section is also known as the ordinal
  $\omega^2$. We identify any pair $(i,j)\in \mathbb{N}\times\mathbb{N}$ with the ordinal $\omega\times i + j$. Thus the three kinds of ordinals in $\omega^2$
  are represented as follows:
  \begin{description}
  \item[null ordinal] : the pair $(0,0)$
  \item[successor ordinal] : any pair $(i,j)$ where $j>0$
  \item[limit ordinal] : any pair $(i,0)$ where $i>0$
  \end{description}
\end{remark}

The detailed  proof script is in the file \url{../src/html/hydras.Hydra.Omega2_Small.html}.

\subsubsection{Preliminaries}
Let us assume there is a variant from \texttt{Hydra} into \texttt{nat*nat} (with the
   lexicographic ordering)  for proving the   termination of all hydra battles.

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Omega2_Small.html}{Hydra.Omega2\_Small}}

\begin{Coqsrc}
  Section Impossibility_Proof.

  Let t := (nat * nat)%type.

  (** non-dependent lexicographic strict ordering on nat*nat *)

  Let lt2 : relation t := lexico Peano.lt Peano.lt.

  Infix "<" := lt2.

  (** reflexive closure of lt2 *)
  Let le2 := clos_refl _ lt2.
  Infix "<=" := le2.

  Variable m : Hydra -> t.

  Context (Hvar : Hvariant lt2_wf free m).
\end{Coqsrc}

Let us follow the same pattern as in Sect.~\ref{omega-case}.
First, we define an injection from type \texttt{t} into \texttt{Hydra}.
We associate with any pair $(i,j)$ the hydra with $i$ branches of length $2$ and
$j$ branches of length $1$.

%% revenir ici

\vspace{4pt}
\emph{From Module ~\href{../src/html/hydras.Hydra.Omega2_Small.html\#iota}{Hydra.Omega2\_Small}}

\begin{Coqsrc}
  Let iota (p: t) := 
    node (hcons_mult (hyd1 head) (fst p)
                     (hcons_mult head (snd p) hnil)).
\end{Coqsrc}

For instance, Figure~\vref{fig:essai2} shows the hydra associated to the pair
$(3,5)$.

\begin{figure}[htb]
\centering
\begin{tikzpicture}[very thick, scale=0.4]
\node (foot) at (6,0) {$\bullet$};
\node (N1) at (1,3) {$\bullet$};
\node (N2) at (3,3) {$\bullet$};
\node (N3) at (5,3) {$\bullet$};
\node (N4) at (8,3) {$\Smiley[2][green]$};
\node (N5) at (11,3) {$\Smiley[2][green]$};
\node (N6) at (14,3) {$\Smiley[2][green]$};
\node (N7) at (17,3){$\Smiley[2][green]$};
\node (N8) at (20,3){$\Smiley[2][green]$};
\node  (N9) at (0,5) {$\Smiley[2][green]$};
\node (N10) at (2,5) {$\Smiley[2][green]$};
\node (N11) at (4,5) {$\Smiley[2][green]$};
\draw (foot) to [bend left=10] (N1);
\draw (foot) -- (N2);
\draw (foot) -- (N3);
\draw (foot) -- (N4);
\draw (foot) -- (N5);
\draw (foot) -- (N6);
\draw (foot) to [bend right=10] (N7);
\draw (foot) to [bend right=15] (N8);
\draw (N1) to [bend left=10] (N9);
\draw (N2) -- (N10);
\draw (N3) -- (N11);
\end{tikzpicture}
\caption{\label{fig:essai2}
The hydra $\iota(3,5)$}
\end{figure}

Like in Sect.~\ref{omega-case}, we build a hydra out of the range of \texttt{iota} (represented in Fig.~\vref{fig:h-omega2-small}).

\begin{figure}[htb]
\centering
\begin{tikzpicture}[very thick, scale=0.5]
\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (3,4) {$\Smiley[2][green]$};
\node (N3) at (1,4) {$\Smiley[2][green]$};
\draw (foot) -- (N1);
\draw (N1) to [bend right =15] (N2);
\draw (N1) to  [bend left=15](N3);
\end{tikzpicture}
\caption{\label{fig:h-omega2-small}}
 The hydra \texttt{big\_h}.
\end{figure}
\begin{Coqsrc}
   Let big_h := hyd1 (hyd2 head head).  
 \end{Coqsrc}

 In a second step, we build a ``smaller'' hydra.

\begin{Coqsrc}
   Let small_h := iota (m big_h).
\end{Coqsrc}

Like in Sect.~\ref{omega-case}, we prove the double inequality \texttt{m big\_h <= m small\_h < m big\_h}, which is impossible.

\subsubsection{Proof of the inequality \texttt{m small\_h < m big\_h}}

For proving the inequality  \texttt{m\_lt: m small\_h < m big\_h}, it suffices to
build a fight transforming \texttt{big\_h} into \texttt{small\_h}.

First we prove that \texttt{small\_h} is reachable from \texttt{big\_h} in one or two steps. Let us decompose \texttt{m big\_h}
into the pair $(i,j)$.
If $j=0$, then one round suffices to transform \texttt{big\_h} into $\iota(i,j)$.
If $j>0$, then a first round transforms \texttt{big\_h} into $\iota(i+1,0)$ and a second round into $\iota(i,j)$. So, we have the folowing result.

\begin{Coqsrc}
 Lemma big_to_small: big_h -+-> small_h.
\end{Coqsrc}

Since $m$ is a variant, we infer the following inequality:

\begin{Coqsrc}
Corollary m_lt : m small_h < m big_h.
\end{Coqsrc}

\subsubsection{Proof of the inequality \texttt{m big\_h <= m small\_h} }

The proof of the inequality \texttt{m big\_h <= m small\_h} is quite more complex than in Sect~\ref{omega-case}.  If we consider some pair $(i,j)$, where $i>0$, there exists an infinite number of
pairs stricly less than $(i,j)$, and there exists an infinite number of battles that start from
$\iota(i,j)$. In effect, at any configuration $\iota(k,0)$, the hydra can freely chose any replication number. Intuitively, the measure of such a hydra must be large enough for taking into account
all the possible battles issued from that hydra.
Let us give more technical details.

\begin{itemize}
\item The proof of the lemma \texttt{m\_ge : forall p : t,   p <= m (iota p)} uses well-founded induction on $p$, and not structural induction on natural numbers

\item For any pair $p$, we have to distinguish between three cases, according to the value of $p$'s components.
  \begin{itemize}
  \item $p=(0,0)$
  \item $p=(i,0)$, where $i>0$\,: $p$ corresponds to a limit ordinal
  \item $p=(i,j)$, where $j>0$\,: $p$ is the successor of $(i,j-1)$.
  \end{itemize}
\end{itemize}

Before starting the proof, we have to express the notion of \emph{limit} in terms of
least upper bounds, through the following logical equivalence.

\vspace{4pt}
\emph{From Module ~\href{../src/html/hydras.Hydra.Omega2_Small.html\#limit_is_lub}{Hydra.Omega2\_Small.v}}.

\begin{Coqsrc}
Lemma limit_is_lub : forall i p, 
   (forall j, (i,j) < p) <-> (S i, 0) <= p.  
\end{Coqsrc}

Let us define the notion of elementary ``step'' of decreasing sequences in
\texttt{t}

\begin{Coqsrc}
Inductive step : t -> t -> Prop :=
| succ_step : forall i j,  step (i, S j) (i, j)
| limit_step : forall i j, step (S i, 0) (i, j).
\end{Coqsrc}

The following lemma establishes a correspondance between the relation
\texttt{step} and hydra fights.

\begin{Coqsrc}
Lemma step_to_fight : forall p q, step p q -> iota p -+-> iota q.
\end{Coqsrc}

\index{Maths!Transfinite induction}

Thus, starting from any inequality $q < p$ on type \texttt{t}, we can build 
by transfinite induction over \texttt{p} a fight 
that transforms the hydra $\iota(p)$ into $\iota(q)$.

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Omega2_Small.html\#m_ge}{Hydra.Omega2\_Small}}

\begin{Coqsrc}
Lemma m_ge : forall p : t,   p <= m (iota p).
Proof.
  intro p ; pattern p;
    apply  well_founded_induction with 
               (R := lt2) (1:= wf_lexico lt_wf lt_wf);
     intros (i,j) IHij (* rest of proof skipped *)
\end{Coqsrc}

\begin{Coqanswer}
  i, j : nat
  IHij : forall y : t, y < (i, j) -> y <= m (iota y)
  ============================
  (i, j) <= m (iota (i, j)) 
\end{Coqanswer}

Then we have  three cases to consider, according to the values of $i$ and $j$.
\begin{itemize}
\item If $p=(0,0)$ then obviously, $\iota(p)\geq p = (0,0)$
\item If  $p=(i+1,0)$ for some $i\in\mathbb{N}$, we
 remark  that $p$ is strictly greater than any pair $ (i, j)$, where $j$ 
is any natural number.

Applying the battle rules, for any $j$, we have $\iota(i+1,j)  {\round} \iota(i, j) $, thus $m(\iota(p)) > m(\iota(i,j)$ since  $m$ is assumed to be a variant.

Applying the induction hypothesis, we get the inequality
 $ m(\iota(i,j)) \geq (i,j)$ for any $j$.

Thus, $m(\iota(p)) > (i,j)$ for any $j$.
Applying the lemma \texttt{limit\_is\_lub}, we get  the inequality
$\iota(i+1,0)\geq (i+1,0)$

\item If $p=(i,j+1)$ with $j\in\mathbb{N}$, we have  $\iota(p)  {\round} \iota(i, j) $,
hence $m(\iota(p))> m(\iota(i,j)) \geq (i,j)$, thus $m(\iota(p))\geq (i,j+1)=p$

\end{itemize}

\subsubsection{End of the proof}

Since $<$ is a strict order (irreflexive  and transitive) on \texttt{nat*nat}, we can conclude that there is no
variant for termination on the lexicographic square of $(\mathbb{N},<)$.

\begin{enumerate}
\item From \texttt{m\_lt}, we infer the strict inequality 
\texttt{m small\_h < m big\_h}
\item from \texttt{m\_ge}, we get \texttt{m big\_h <= m (iota (m big\_h)) = m small\_h} 
\end{enumerate}

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Omega2_Small.html}{Hydra.Omega2\_Small}}

\begin{Coqsrc}
Theorem Impossible : False.
Proof.
  destruct (StrictOrder_Irreflexive  (m big_h)).
  apply le2_lt2_trans with (m small_h).
  -  unfold small_h;  apply m_ge.
  -  apply m_lt. 
Qed.

End Impossibility_Proof.
\end{Coqsrc}

\begin{exercise}
Prove that there exists no variant $m$ from \texttt{Hydra} into \texttt{nat*nat} for proving
    the  termination of all \emph{standard} battles.
\end{exercise}

\begin{remark}
In Chapter~\ref{ks-chapter}, we will prove a much more general  theorem
than the two previous results about $\mathbb{N}$ and $\mathbb{N}\times\mathbb{N}$. The proof of that general  result will share the same structure, but will 
require a lot of technical results. 
\end{remark}

\begin{exercise}

\label{sec:orgheadline63}
Write \emph{direct} proofs ({i.e.},  without applying the result and tools of Chap.~\ref{ks-chapter}) that the following data structures  are too simple for defining a variant for any hydra battle.

\begin{itemize}
\item  $\omega^n$ : the set of all $n$-uples of natural numbers, ordered  by 
  lexicographic ordering
\item  $\omega^\omega$: the set of all decreasing sequences (with respect to $\le$)  of natural numbers, ordered by lexicographic ordering on lists.

For instance , the sequence $\langle 4,4,2 \rangle$ is strictly greater than 
$\langle 4,3,3,3,3,3,3,2,2,2 \rangle$.
\end{itemize}

\end{exercise}
